Bayesian exponential random graph modeling of KWL Burial network

### 2020-10-28 analysis/paper/paper.Rmd
adjust gamma and add orientation variable

<!--burial study- network analysis-->
Burials contain valuable information about both individual mortuary practices and the structure of a society. The development of network science and graph theory provides many new tools for the study of the relations between burials and exploration their social implications [@Sosna2013]. Mortuary practices such as burial forms, grave goods, and related ritual practices that structure the material configuration of burials can represent social relations between members of a community, allowing the inference for social structures [citation needed]. This means we can use data from burials to study social differences to answer questions related to social inequality and complexity [@Ehrhardt2009; @Gamble2002; @Janes2013; @Johansen2015]. For example, burial goods can reflect social practices in broader cultural contexts to represent personal wealth, social roles, or traditional customs. Burials with grave goods in common often indicate some similar underlying social relations, for example, individuals that share similar access to trade, exchange, and gifting networks [@Coward2013, pp. 252]. Identifying the structure of relationships among burials can give insights into the structure of social organization.

@@

<!--current approaches to network analysis-->
Exponential Random Graph Models (ERGMs) are an important family of statistical models for networks that allows directly modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. ERGMs differs from traditional network approaches that edges are only formed according to their corresponding predictors separately without taken into account the interactions among edges [@Morris2008]. In ERGMs, networks are viewed as dependent variables, the formation of a tie is also influenced by network dependencies, and the attributes of nodes/edges [@Snijders2011]. For example, nodes with similar attributes is more likely to form a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), or popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics. We can model those network statistics as direct functions of ties, where the probability of occurrence of ties can be specified and hypothesized [@Morris2008; @Robins2007]. By specifying the forms of configurations, we can build a hypothesis-based model to generate a distribution of random networks. Such distribution consists of a large number of possible networks that provides a statistical inference for an observed network [@Robins2007].  This helps to understand whether an observed network shows significantly more or less property of interest than the random networks we modeled.

@@

<!--Network construction and specification-->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relationships among burials and test our hypothesis of social change using the R programming language with bergm package [@Caimo2014], which built upon ergm package for simulation and model specification [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with archaeological interpretations. Every parameter in an ERGM has an associated algorithm for computing its value for a network [@Morris2008]. Based on our hypothesis, we model a network with increased social inequality that represented by low reciprocity, low transitivity, high popularity, high age homophily, and less proximity structure effects. We modeled several node attributes for homophily effects, including age, gender, ritual, and the degree of wealth, to explore the correlation between the tendency of connected nodes and the network ties overall. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried close to each other. Our model infers the presence of a few individuals as centers having more relations with others, which hints the emergence of social inequality. We expect that observed network after the European arrival will have a better match with our model compared with the network before European contact.

<!--Model fitting and prior setting--> 
After the specification, the network is simulated in a Baysesian framework using Markov chain Monte Carlo (MCMC) algorithm. MCMC algorithm is an estimation of posterior distribution of interest through directly random sampling the posterior without assuming the prior is a normal distribution. We can obtain a distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on algorithms. This enables a random or stochastic simulation in a long run without depending on the previous move. More chains ensures a more desirable distribution close to the target distribution that we study, which is called the convergence of the Markov chains. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and than switch to a pair of actors at random within the chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000.

<!-- Goodness-of-fit assessment-->
The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. This provides a statistical approach to check how well the estimated posterior parameter distribution can reproduce networks similar to the general structural features of the observed network. Our observed networks, pre-European and post-European networks, followed the normal distribution that is a suitable prior model in network analysis that assumes networks with low density and high transitivity commonly observed in real world [@Caimo2017].

>>>>

<!--burial study- network analysis-->
Burials contain valuable information about both individual mortuary practices and the structure of a society . The development of network science and graph theory provides many new tools for the study of the relations between burials and exploration their social implications [@Sosna2013]. Mortuary practices such as burial forms, grave goods, and related ritual practices that structure the material configuration of burials can represent social relations between members of a community, allowing the inference for social structures [citation needed]. This means we can use data from burials to study social differences to answer questions related to social inequality and complexity [@Ehrhardt2009; @Gamble2002; @Janes2013; @Johansen2015]. For example, burial goods can reflect social practices in broader cultural contexts to represent personal wealth, social roles, or traditional customs. Burials with grave goods in common often indicate some similar underlying social relations, for example, individuals that share similar access to trade, exchange, and gifting networks [@Coward2013, pp. 252]. Identifying the structure of relationships among burials can give insights into the structure of social organization.

@@

<!--current approaches to network analysis-->
Exponential Random Graph Models (ERGMs) are an important family of statistical models for networks that allows directly modeling for the formation of edges, or ties, between nodes [@Robins2007]. The assumption is that possible ties in a network are random variables and dependent on actor variables or the presence or absence of other ties [@Robins2007]. ERGMs differs from traditional network approaches that edges are only formed according to their corresponding predictors separately without taken into account the interactions among edges [@Morris2008]. In ERGMs, networks are viewed as dependent variables, the formation of a tie is also influenced by network dependencies, and the attributes of nodes/edges [@Snijders2011]. For example, nodes with similar attributes is more likely to form a relationship, such as friendship between people with the same hobby. Ties form a small structure in a network called a graph configuration that describes the form of dependence, such as reciprocity (relationship between two actors), transitivity or clustering (relationship between two actors through a shared third actor), homophily (relationship between actors with a similar attribute), or popularity (actors have many relationships with others) [@Robins2007; @Snijders2006; @Morris2008]. Those configurations represent the structure or the property of a network and can be expressed by network statistics. We can model those network statistics as direct functions of ties, where the probability of occurrence of ties can be specified and hypothesized [@Morris2008; @Robins2007]. By specifying the forms of configurations, we can build a hypothesis-based model to generate a distribution of random networks. Such distribution consists of a large number of possible networks that provides a statistical inference for an observed network [@Robins2007].  This helps to understand whether an observed network shows significantly more or less property of interest than the random networks we modeled.

@@

<!--Network construction and specification-->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language with the bergm package [@Caimo2014], which built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with archaeological interpretations. Every parameter in an ERGM has an associated algorithm for computing its value for a network [@Morris2008]. Based on our hypothesis, we model a network with increased social inequality that represented by low reciprocity, low transitivity, high popularity, high age homophily, and less proximity structure effects. We modeled several node attributes for homophily effects, including age, gender, ritual, and the degree of wealth, to explore the correlation between the tendency of connected nodes and the network ties overall. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried close to each other. Our model infers the presence of a few individuals as centers having more relations with others, which hints the emergence of social inequality. We expect that observed network after the European arrival will have a better match with our model compared with the network before European contact.

<!--Model fitting and prior setting--> 
After the specification, the network is simulated in a Baysesian framework using Markov chain Monte Carlo (MCMC) algorithm. MCMC algorithm is an estimation of posterior distribution of interest through directly random sampling the posterior without assuming the prior is a normal distribution. We can obtain a distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on algorithms. This enables a random or stochastic simulation in a long run without depending on the previous move. More chains ensures a more desirable distribution close to the target distribution that we study, which is called the convergence of the Markov chains. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and than switch to a pair of actors at random within the chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000. For prior, the normal distribution is agreed as a suitable prior model in network analysis that assumes networks with low density and high transitivity commonly found in real world [@Caimo2017]. To

<!-- Goodness-of-fit assessment-->
The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. The Bayesian GOF summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution based on our hypotheses can reproduce networks similar to the general structural features of the observed network. We then compared the distribution of our observed networks, pre-European and post-European networks, with the distribution of our hypothesized model with higher popularity. We expect to see the network after the European presence has a better fit with the hypothesis, indicative of a structure with higher social differentiation.
